---
title: "SpatialOncoSimul with OncoSimulR"
author: "Antonio Giráldez Trujillo, Mercedes Núñez Bayón, María González Bermejo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SpatialOncoSimul function}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

SpatialOncoSimul is a function implemented inside the package OncoSimulR for the
spatial simulation of tumoral cell growth. This function will simulate tumor 
progression from one starting position in a 1D, 2D or 3D grid specified by the user. 
After the first population of tumoral cells is developed with OncoSimulIndiv, a 
small population of tumoral cells will migrate to adjacent or remote spaces based 
on the fitness of the tumor genotypes. The migrating cells will occupied new space 
coordinates and will be combined with an initial population of wild-type cells
present in the new location. Then, the OncoSimulIndiv function will be launched 
again in each space coordinate to simulate tumor progression from the new tumor 
composition in each position. This process is repeated the number of times specified 
by the user to analyse the tumor progression in a spatial context.

The algorithm of SpatialOncoSimul is based on the main concepts of different spatial
models (Bartlomiej Waclaw, et al., 2015, A. Sottoriva, et al., 2015, R. Sun, et al., 
2017, R. Sun, et al., 2021, Noble, R. et al., 2022) and the ideas set out in the 
Master's degree final project of Alberto Parramón Castillo (2018).



Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

Using this package will often involve the following steps:

1. Specify fitness effects.

2. Simulate cancer progression in a spatial context. 

3. Decide on a model. This basically amounts to choosing a model with exponential 
growth (“Exp” or “Bozic”) or a model with carrying capacity (“McFL”). If exponential 
growth, you can choose whether the effects of mutations operate on the death rate 
(“Bozic”) or the birth rate (“Exp”). It is recommended to use the model with
carrying capacity ("McFL") as the population in this model is kept more constant
during the simulation, which biologically is more meaningful as the cell growth 
inside the grid is limited by space. Whereas in the "Exp" o "Bozic" cell population 
can growth disproportionally or tend to 0. 

4. You must also specify the number of iterations for running the simulation and 
the spatial model for representing cell migration.

Of course, at least for initial playing around, you can use the defaults.



```{r, fig.show='hold'}
## Here we specified the Fitness Effects of the population genotypes with restrictions 
# in the order of accumulation of mutations using a DAG.
s1 <- allFitnessEffects(
  data.frame(parent = c("Root", "Root", "i"),
             child = c("u" , "i" , "v"),
             s = c(0.1 , -0.05 , 0.25),
             sh = -1,
             typeDep = "MN"),
  epistasis = c("u:i" = -1,"u:v" = -1))

evalAllGenotypes (s1 , order = FALSE, addwt = TRUE)

# Simul the spatial tumor cell growth in 3D during 20 iterations.
Spatial_3D <- SpatialOncoSimul(fp = s1,
                      model = "McFL",
                      onlyCancer = FALSE,
                      finalTime = 500,
                      mu = 1e-4,
                      initSize = 1000,
                      keepPhylog = FALSE,
                      seed = NULL,
                      errorHitMaxTries = FALSE,
                      errorHitWallTime = FALSE, initMutant = c("i"), 
                      spatialIterMax = 20, SpatialModel = "3D")
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
# We define fitness effects with a new genotype composition.
pancr <- allFitnessEffects(
    data.frame(parent = c("Root", rep("KRAS", 4), 
                   "SMAD4", "CDNK2A", 
                   "TP53", "TP53", "MLL3"),
               child = c("KRAS","SMAD4", "CDNK2A", 
                   "TP53", "MLL3",
                   rep("PXDN", 3), rep("TGFBR2", 2)),
               s = 0.1,
               sh = -0.9,
               typeDep = "MN"),
    drvNames = c("KRAS", "SMAD4", "CDNK2A", "TP53", 
                 "MLL3", "TGFBR2", "PXDN"))
evalAllGenotypes (pancr , order = FALSE, addwt = TRUE)

Spatial_3D <- SpatialOncoSimul(fp = pancr,
                      model = "McFL",
                      onlyCancer = FALSE,
                      finalTime = 500,
                      mu = 1e-4,
                      initSize = 1000,
                      keepPhylog = FALSE,
                      seed = NULL,
                      errorHitMaxTries = FALSE,
                      errorHitWallTime = FALSE, 
                      spatialIterMax = 10, SpatialModel = "3D")
```
## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
# We define fitness effects with a new genotype composition.
s2 <- allFitnessEffects(epistasis = c ("A:-B:-C"=0.05, "-A:B:-C"=0.05 , 
                                       "-A:-B:C"=0.05 ,"A:B:-C"=0.045 , 
                                       "A:-B:C"=0.045 , "-A:B:C"=0.045 ,
                                       "A:B:C"=0.055))
                                      
# Simulation of spatial model.
sp2 <- SpatialOncoSimul (s2, model = 'McFL',
                         onlyCancer = FALSE,
                         finalTime = 500,
                         mu = 1e-4,
                         initSize = 1000,
                         initMutant = c ("A"),
                         spatialIterMax = 10, SpatialModel = "3D")



```
Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
