---
title: "SpatialOncoSimul with OncoSimulR"
author: "Antonio Giráldez Trujillo, Mercedes Núñez Bayón, María González Bermejo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SpatialOncoSimul function}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

SpatialOncoSimul is a function implemented inside the package OncoSimulR for the
spatial simulation of tumoral cell growth. This function will simulate tumor 
progression from one starting position in a 1D, 2D or 3D grid specified by the user. 
After the first population of tumoral cells is developed with OncoSimulIndiv, a 
small population of tumoral cells will migrate to adjacent or remote spaces based 
on the fitness of the tumor genotypes. The migrating cells will occupied new space 
coordinates and will be combined with an initial population of wild-type cells
present in the new location. Then, the OncoSimulIndiv function will be launched 
again in each space coordinate to simulate tumor progression from the new tumor 
composition in each position. This process is repeated the number of times specified 
by the user to analyse the tumor progression in a spatial context.

The algorithm of SpatialOncoSimul is based on the main concepts of different spatial
models (Bartlomiej Waclaw, et al., 2015, A. Sottoriva, et al., 2015, R. Sun, et al., 
2017, R. Sun, et al., 2021, Noble, R. et al., 2022) and the ideas set out in the 
Master's degree final project of Alberto Parramón Castillo (2018).



## Vignette Info

Using this function will often involve the following steps:

1. Specify fitness effects.


2. Simulate cancer progression in a spatial context.Decide on a model. 
This basically amounts to choosing a model with exponential growth (“Exp” or “Bozic”) 
or a model with carrying capacity (“McFL”). If exponential growth, you can choose 
whether the effects of mutations operate on the death rate (“Bozic”) or the birth 
rate (“Exp”). It is recommended to use the model with carrying capacity ("McFL") 
as the population in this model is kept more constant during the simulation, which 
biologically is more meaningful as the cell growth inside the grid is limited by 
space. Whereas in the "Exp" o "Bozic" cell population can grow indefinitely or 
decrease until extinct.

3. You must also specify the number of iterations for running the simulation and 
the spatial model for representing cell migration. The percentage of migrating cells,
the probability of migration to near and remote spaces can also be specified by 
the user.

Of course, at least for initial playing around, you can use the defaults.



```{r, fig.show='hold'}
## Here we specified the Fitness Effects of the population genotypes with restrictions 
# in the order of accumulation of mutations using a DAG.
s1 <- allFitnessEffects(
  data.frame(parent = c("Root", "Root", "i"),
             child = c("u" , "i" , "v"),
             s = c(0.1 , -0.05 , 0.25),
             sh = -1,
             typeDep = "MN"),
  epistasis = c("u:i" = -1,"u:v" = -1))

evalAllGenotypes (s1 , order = FALSE, addwt = TRUE)

# Simul the spatial tumor cell growth in 3D during 20 iterations, using default
# migration parameters  migrationProb = 0.5, largeDistMigrationProb = 1e-6, 
# maxMigrationPercentage = 0.2

Spatial_3D <- SpatialOncoSimul(fp = s1,
                      model = "McFL",
                      onlyCancer = FALSE,
                      finalTime = 500,
                      mu = 1e-4,
                      initSize = 1000,
                      keepPhylog = FALSE,
                      seed = NULL,
                      errorHitMaxTries = FALSE,
                      errorHitWallTime = FALSE, initMutant = c("i"), 
                      spatialIterMax = 20, SpatialModel = "3D")
```


## As a second example, we will use a model where we specify restrictions in the 
## order of accumulation of mutations using a DAG with the pancreatic cancer poset 
## in Gerstung, Eriksson, et al. (2011).


```{r, echo=FALSE, results='asis'}
# We define fitness effects with a new genotype composition.
pancr <- allFitnessEffects(
    data.frame(parent = c("Root", rep("KRAS", 4), 
                   "SMAD4", "CDNK2A", 
                   "TP53", "TP53", "MLL3"),
               child = c("KRAS","SMAD4", "CDNK2A", 
                   "TP53", "MLL3",
                   rep("PXDN", 3), rep("TGFBR2", 2)),
               s = 0.1,
               sh = -0.9,
               typeDep = "MN"),
    drvNames = c("KRAS", "SMAD4", "CDNK2A", "TP53", 
                 "MLL3", "TGFBR2", "PXDN"))
evalAllGenotypes (pancr , order = FALSE, addwt = TRUE)

Spatial_3D <- SpatialOncoSimul(fp = pancr,
                      model = "McFL",
                      onlyCancer = FALSE,
                      finalTime = 500,
                      mu = 1e-4,
                      initSize = 1000,
                      keepPhylog = FALSE,
                      seed = NULL,
                      errorHitMaxTries = FALSE,
                      errorHitWallTime = FALSE, 
                      spatialIterMax = 10, SpatialModel = "3D")
```

## Testing the SpatialOncoSimul function using functionality of frequency-dependent 
## fitness in OncoSimulR package. 

With frequency-dependence fitness we can make fitness (actually, birth rate) depend 
on the frequency of other genotypes. We specify how the fitness (birth rate) of 
each genotype depends (or not) on other genotypes.


```{r, echo=FALSE, results='asis'}
## Define fitness of the different genotypes
gffd <- data.frame(
    Genotype = c("WT", "A", "B", "C", "A, B"), 
    Fitness = c("1 + 1.5 * f_A_B",
                "1.3 + 1.5 * f_A_B",
                "1.4",
                "1.1 + 0.7*((f_A + f_B) > 0.3) + f_A_B",
                "1.2 + sqrt(f_1 + f_C + f_B) - 0.3 * (f_A_B > 0.5)"))

afd <- allFitnessEffects(genotFitness = gffd, 
                         frequencyDependentFitness = TRUE,
                         frequencyType = "rel")
set.seed(1) ## for reproducibility
sfd <- SpatialOncoSimul(afd, 
                       model = 'McFL',
                       onlyCancer = FALSE,
                       finalTime = 55, ## short, for speed
                       mu = 1e-4,
                       initSize = 1000,
                       initMutant = c ("A"),
                       spatialIterMax = 10, SpatialModel = "3D")
```


## As a fourth example, to hightly the importance of migrationProb parameter, 
## a simulation is runned with the maximum probability of migration to near spaces.

```{r, echo=FALSE, results='asis'}
# We define fitness effects with a new genotype composition.
s2 <- allFitnessEffects(epistasis = c ("A:-B:-C"=0.05, "-A:B:-C"=0.05 , 
                                       "-A:-B:C"=0.05 ,"A:B:-C"=0.045 , 
                                       "A:-B:C"=0.045 , "-A:B:C"=0.045 ,
                                       "A:B:C"=0.055))
                                   
# Simulation of spatial model changing the migration probability to near spaces.
sp2 <- SpatialOncoSimul (s2, model = 'McFL',
                         onlyCancer = FALSE,
                         finalTime = 500,
                         mu = 1e-4,
                         initSize = 1000,
                         initMutant = c ("A"),
                         migrationProb = 1, 
                         largeDistMigrationProb = 1e-6, 
                         maxMigrationPercentage = 0.2,
                         spatialIterMax = 10, SpatialModel = "2D")



```
Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
